---
title: "RegressionCars"
author: "Lindsey Erickson"
date: "May 22, 2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

#Introduction
This assignment takes the cars dataset located in the datasets library and explores the relationship between a set of variables and miles per gallon (MPG).  We're particularly interested in the MPG differences between automatic transmission or manual transmissions, and which transmission type is better for MPG.

#Limitations
This analysis is for an assignment that asks to do some quick regression building to solve the above questions.  Given more time, a deeper analysis and better model building techniques.

#Executive Summary
I did some exploratory analysis and model building to understand the miles per gallon (MPG) on different cars.  Looking at the data anaylsis, it was clear that cars with a manual transmission had a higher MPG than cars with an automatic transmission.  I ran a t-test to see if there was any significance in the previous comment, which there was.  Therefore, I knew that manual tranmission cars had a higher MPG than automatic transmission. Because of this, I knew I needed to include transmission type when building a regression model to determine what kind of car has the highest MPG.

I started the regression model building by first running a correlation plot to see what variables to include in the regression model.  I knew from the above comments that I wanted to include transmission type; however, I needed the correlation plot to see what else to include.  Out of curiosity, I ran a model that included every available variable.  This model was not ideal; however, it produced an R-sqaure of about 0.87, the residual vs fitted plot looked random, and the QQ-plot shows that the residuals are fairly normal.  This model appears to be really good; however, I suspect it would be an overfitted model.  However, running a model with the top seven correlated variables produced the most random (no patterns) residual vs fitted plot and normal QQ-plot.  

#Final Model
The final model used was mpg = factor(cyl) + disp + drat + factor(am) + wt + factor(vs) + hp, which produced an R-sqaured of >0.87 and decent residual plots.

#Exploratory Analysis
To do this, I"ll first do some exploratory analysis to make some sense of the data.  These are the steps I'll take:

    1. look at the first six observations and the last six observations
    2. look at the structure of the data
    3. look some summary statistics
    4. check for missing data
    5. run correlation plots (see appendix)
    
```{r, echo=TRUE, collapse=TRUE, warning=FALSE}
library(datasets)
library(ggplot2)
library("PerformanceAnalytics")

dataCars <- mtcars
head(dataCars, n = 2) #first two observations
tail(dataCars, n = 2) #last two observations
summary(dataCars) #summary of the data
str(dataCars) #structure of the data

#checking for missing data
table(is.na(dataCars))

#checking data point that looks like an outlier
dataCars[ which(dataCars$hp == 335.0), ]

```
</br>
Here's what we know for running the above analysis:
  
  1. There are 32 observations and 11 variables
  2. All variables are numeric
  3. None of the variables seem to have any outliers; however, there is a value of 335 in hp.  This seemed like an outlier to me; however, when checking the value, I noticed that it was associated with a Maserati so I will leave it as is.
  4. There is no missing values in any of the variables
</br>

Since we are interested in the MPG between the two different transmission types, let's take a look at a box plot to see the variation of MPG between automatic transmission and manual transmission.

```{r, echo=TRUE, collapse=TRUE}
cor(dataCars$am, dataCars$mpg) #correlation between transmission and MPG
t.test(dataCars$mpg~dataCars$am) #ttest for statistical significance between transmission types and MPG
pValue <- round(t.test(dataCars$mpg~dataCars$am)$p.value, 3)

ggplot(dataCars, aes(x=as.factor(am), y=mpg)) +
  geom_boxplot(fill="hot pink") +
  xlab("Transmission Type") +
  ylab("MPG") +
  ggtitle("Boxplot of mean miles per gallon (MPG) by transmission type") +
  scale_x_discrete(labels=c("Automatic","Manual"))
```
</br>
First, we see that the correlation between MPG and transmission type is 0.60.  Not bad.  Looking at the boxplot, visually, it seems that manual transmission provides a higher MPG than automatic transmission.  The t-test shows that there is a statistical difference in the MPG between automatic transmissions and manual transmissions (p-value: `r pValue`)
</br>

#Fit our Regression Model
Now, we will fit the models.  The first model uses all variables, the second model uses the top three correlated variables.  The last model uses top seven correlated variables.
```{r, echo=TRUE, collapse=TRUE}
model1 <- lm(mpg ~ ., data=dataCars)
summary(model1)$r.squared

model2 <- lm(mpg ~ factor(cyl) + disp + wt, data=dataCars)
summary(model2)$r.squared

model3 <- lm(mpg ~ factor(cyl) + disp + drat + factor(am) + wt + factor(vs) + hp, data=dataCars)
summary(model3)

```
</br>

#Appedix

correlation matrix on the dataset
```{r, echo=TRUE, collapse=TRUE, warning=FALSE}
#my_data <- dataCars[, c(1,3,4,5,6,7, 8, 9, 10, 11)]
chart.Correlation(dataCars, histogram=TRUE, pch=19)
```
</br>

diagnostic plots for model3: residual vs fitted plot and QQ-Plot to see whether the residuals are normal
```{r, echo=TRUE, collapse=TRUE, warning=FALSE}
plot(model3, which = 1)  #see residuals vs fitted
plot(model3, which = 2) #see whether the residuals are normal
```
